Investigating the impact of interventionist causal approach on the study of verbal aggression online

### 2021-09-09 paperDraft/paperDraft.Rmd
changes made to philosophical grounding of intevrentions

---
title: "Investigating the impact of interventionist causal approach on the study of verbal aggression and discrimination online"
author: "Patrycja Tempska and Rafal Urbaniak"
output:
  pdf_document:
    number_sections: yes
    df_print: kable
    keep_tex: yes
    includes:
      in_header: Rafal_latex6.sty
  html_document:
    df_print: paged
  word_document: default
classoption: dvipsnames,enabledeprecatedfontcommands
fontsize: 10pt
documentclass: scrartcl
urlcolor: blue
bibliography: ../references/attacks.bib
csl: ../references/apa-6th-edition.csl
---

@@

# Section Abstract 
This article \todo{hm not sure whether abstract is neccessary in BA thesis but I will leave it here for now} describes an experimental intervention study conducted in a naturalistic, digital setting (Q&A forum on Reddit), utilizing a collective intelligence approach to content moderation and reduction of the level of verbal aggression among a selected group of Reddit users who regularly attack other community members. Collective Intelligence in this sense means exploring the collaboration between human and machine intelligence to develop solutions to social challenges. Artificial Intelligence was used to detect verbal aggression (personal attacks) and notify human volunteers about attacks. Volunteers after receiving notifications employed interventions based on norm or empathy promotion. We find that only those who were sanctioned with norms-inducing interventions had their personal attacks' level significantly decreased.

@@

We live in a world in which there is an abundance of data. Paraphrasing the phrase coined by Clive Humby, "data is the new oil" --- data has displaced oil and become the world’s most valuable resource. Albeit there are some fundamental differences between the two assets. Oil belongs to a very limited and finite resource, while data is growing at an exponential rate (cite: https://cloudtweaks.com/2015/03/how-much-data-is-produced-every-day/). The value of oil does not depend on its amount or the refinery processes - it's rather constant at a particular dynamic of market's supply and demand. Data and its value, on the contrary, is inherently tied to its amount and the "refinery process" - various tools and techniques for its discovery and analysis. In some cases, certain techniques can be used only if one has a sufficient (meaning enormous) data at one's disposal --- e.g. statistical methods like machine learning and deep learning \todo{@article{lecun2015deep,

@@

In the last couple of years, more and more researchers started to research and developed tools for causal discovery and inference. Unfortunately, although thinking in causal terms seems to be our inherent human capacity and one of our basic cognitive abilities, that is not necessarily the case with the majority of toolkits and statistical models developed to draw insight and correlations from data. 

@@

But often scientists conduct the so-called cross-sectional research, in which no variable is manipulated. In the language and toolkit of classical statistics in such cases we cannot conduct any causal inference. What we can measure using those methods is correlation, and, as frequently repeated, “correlation does not imply causation”. Correlation can be positive, negative, or there can be no correlation at all. A positive one implies that the increase of the value of one variable will influence the increase of another variables’ value. A negative correlation means that the change in values of two variables will diverge into separate directions --- one will increase and one will decrease alongside the first one. We can measure the strength of the correlation using various mathematical methods, where 1 or -1 means we are dealing with perfect correlation --- the value of one variable changes proportional to the value of another variable. But even though causation is necessarily connected with correlation, as already mentioned --- not every correlation implies causation. And as we will see, ignoring the causal story behind the data and relationships between the variables might lead to far-reaching consequences. Drawing false causal conclusions out of data can lead to false accusations of discrimination (...here I will add something more once I decide which examples to choose). This is especially problematic if we take into consideration the fact that statisticians and people in general naturally think in causal terms and spurious interpretations of studies occur frequently. On the other hand, a conservative approach of many classical statisticians led to a long-living debate on whether smoking causes cancer, lasting for over 20 years (to check). One of the reasons for such a fierce debate was the lack of a causal framework, which was still in its infancy at that time. Framework of tools with which we could validate or invalidate assumptions about the existence of the potential confounder - a vari- able causing both craving cigarettes and lung cancer (e.g. some genetic factor), creating a spurious correlation between cigarettes and lung cancer.
Modeling causality and moving beyond traditional statistics is a relatively young endeavor. One of the pioneers of the work on causality is Judea Pearl. As he describes, causality and traditional statistics are governed by opposing ways of thinking. While statistics might be described as “study of methods of the reduction of data” (quoted here was Fischer I think, need to check), causality requires the incorporation of an extra layer - the understanding or prior knowl- edge (assumptions) about the process that generate the data in the first place. Classical statistics might be described here as a “model-free approach,” just as Artificial Intelligence methods based on machine learning/deep learning, in which no prior knowledge can be taken into account.

In this paper, I will show how mathematical methods for causality modeling can be applied to discover how different variables affect each other, moving beyond traditional correlation. Data used for analysis comes from an experimental intervention study conducted in a naturalistic, digital setting (Q&A forum on Reddit), utilizing a collective intelligence approach to content moderation and reduction of the level of verbal aggression among a selected group of Reddit users who regularly attack other community members. Collective Intelligence in this sense means exploring the collaboration between human and machine intelligence to develop solutions to social challenges. Artificial Intelligence was used to detect verbal aggression (personal attacks) and notify human volunteers about attacks. Volunteers after receiving notifications employed interventions grounded in philosophy. There were two broad categories - one based on norms (deontology, virtue ethics among others) and second based on empathy (inspired by notions of David Hume and Adam Smith).

Although much effort has been made in order to tackle the problem of verbal aggression and harassment online, looking at various reports and surveys [@sorrentino2019epidemiology; @vogels_state_2021; @Zachary_hate], it remains a common hindrance for people engaging with social media in their everyday lives. The problem got exacerbated in the midst of the COVID19 pandemic, during which the majority of our social life moved to cyberspace. During this shift, there was an increase in cyberbullying attitudes and perpetration [@barlett2021comparing], 90% increase in public reports of illegal online content [@grant_2021], including 114% increase in non-consensual sharing of intimate images, 30% increase in cyberbullying, as well as 40% of increase in adults reporting online harassment. According to a report conducted by company L1ght [@noauthor_l1ght_2020], hate speech directed towards China and the Chinese went up by 900% on Twitter. Gaming platforms were in the spotlight as well, with a 40% increase in toxicity on Discord \footnote{In both reports, the increase is reported as a relative change between the year 2019 and 2020, with no absolute indicators.}.

But alongside the growing need for even more efficient and proactive moderation, the capacity to execute it did not go hand in hand, forcing companies and policymakers to rethink the current model of moderation processes and workforce. Due to the COVID19 restrictions including social distancing, a lot of those serving the role of moderators had to be sent home [@bhattacharya] without the ability to work remotely because of the constraints affiliated with restrictive non-disclosure agreements (NDA) among others. Curtailing the moderators' workforce was accompanied by more agency given to algorithms and AI-based moderation. Those changes, as argued by @gerrard2020covid19, can be seen as a serious red flag in terms of safety for all users on online platforms.

@@

# Collective Intelligence Approach to Counter-speech
Traditionally collective intelligence has been defined as "a group or a team’s combined capacity and capability to perform a wide variety of tasks and solve diverse problems" [@noauthor_collective_nodate]. In our paper and in the theoretical underpinnings of the experiment itself, we will be relying on a collective intelligence scope proposed by Nesta, an innovation foundation (https://www.nesta.org.uk), which focuses on a collaboration between human and machine intelligence to develop innovative solutions to social challenges. (Gdzieś stopka(?): Samurai Labs has been one of the 15 recipients of the second round of collective intelligence grants awarded by Nesta: https://www.nesta.org.uk/project-updates/second-round-collective-intelligence-grants/)

@@

\textbf{Empathetic interventions}
In the first treatment condition, volunteers were encouraged to send empathy-inducing messages focusing either on the target of verbal aggression (e.g. „Hey such words might hurt the other person”), stressing the common humanity aspect that we call share („We are all humans of flesh and blood”), or even infusing the intervention with the emphatic response to the attacker (Hey I understand your strong emotions…”). 
At the core of the empathetic interventions, we put forth the notion that goes back to David Hume and Adam Smith. The first one conceived empathy (at the time referred to as sympathy) as mirroring the emotional state of another person. In the academic psychological literature, similar phenomenon was distinguished and coined in the term emotional contagion
(https://www.sciencedirect.com/topics/psychology/emotional-contagion): "the process in which an observed behavioral change in one individual leads to the reflexive production of the same behavior by other individuals in close proximity, with the likely outcome of converging emotionally". For Adam Smith, sympathy consisted of visualizing how the sympathetic person would feel in the particular circumstances of the other --- thus here the process was based not so much on mirroring, but rather projecting my imagination of what it is like to be that person in a certain moment). Without further delving into the differentiation between those two, sympathy in both accounts is crucial in the constitution of human beings as social and moral creatures (https://plato.stanford.edu/entries/empathy/). It enables the emotional connection to others and concern for their well-being. 
In the psychological literature, various kinds of empathetic responses were distinguished - the aforementioned emotional contagion, affective/proper empathy, sympathy, personal distress or cognitive empathy ((https://plato.stanford.edu/entries/empathy/). During the experiment, we couldn’t observe or measure whether the empathetic response indeed was evoked. Also, interventions that stated that "such words might hurt the other person" have the underlying assumption that the receiver of the attack might be hurt, but in reality that might not even be the case. Intuitively, and following the argumentation of Thomas Nagel in “What is it like to be a bat?” one can only imagine what it is like - but for me - to be a bat or to be a receiver of the attack. To each observed or imagined experience there is an array of subjective quality to actually experiencing it and in this way receiving a particular message might be met with a unique reception by each conscious being. Also, full epistemic access to the mind and body of another is impossible. But even though such access is impossible and putting aside the broad spectrum of phenomena related to empathetic response, our goal was by utilizing interventions referring to empathy in various forms, changing the behavior of the attacker - convincing him to refrain from using verbal aggression towards others. Whether the message indeed gave rise to empathy was not of importance - one can imagine a hypothetical scenario of a successful intervention in which the attacker changed his behavior long-term and stopped attacking others, hopefully as a result of interventions, but the  empathetic response was not even evoked. The behavioral change could be induced by other motivations - e.g. unwillingness from the exclusion from the community. Thus the goal is not to evoke empathy, but through empathetic interventions - substantially limiting the use of verbal aggression.

\textbf{Normative interventions}
Normative interventions were grounded in broadly defined norms and normative theories expressed via various means - either specific community guidelines imposed by Reddit or through ethics - deontology, virtue ethics. Whatever the strategy was used, it was supposed to express the unacceptability of verbal aggression in a direct or non-direct way.

A more direct approach referred to deontology in which „actions are good or bad according to a clear set of rules” (https://ethics.org.au/ethics-explainer-deontology/). Such interventions could state that „we have a duty to respect each other during the discussions", „we have a moral obligation to act in a civil manner while participating in online communities”, „we shouldn’t use ad personam in here”. The basic assumption behind those interventions is that there is something we ought to do or are morally required to do - and such actions are the right actions. There are also actions we ought not to do - and such actions are morally wrongful. This line of thinking can be traced back to Immanuel Kant who thought that all universal moral obligations can stem from categorical imperative: “act only in accordance with that maxim through which you can at the same time will that it become a universal law” (Groundwork of the Metaphysic of Morals).

Yet other interventions were encouraged to be expressed in the light of virtue ethics and stressed the importance of adhering to or practicing certain virtues. Such a message was assumed to motivate the moral agent in the process of positive reinforcement and create a feeling of the desirability of certain behaviors, e.g. „capacity to be respectful in a heated discussion is a virtue and requires hard work”. The core assumption here is that the virtue itself is not genetically inherited but is rather a potential or a disposition of a character that can be practice and mastered like a practical skill (even though as mentioned by Natasza Szutta in [cite her book about virtue ethics] virtues and practical skills share some fundamental differences). NS distinguished two types of virtues - affective and cognitive. Affective virtues encompass the emotions and feelings that play an important and positive role in morality and can act as a support in the course of becoming a virtuous man. A cognitive virtue relates to the intellectual aspect in which one knows how to act in certain situations and understands the rules of morally rightful actions (e.g. what is kindness and how a kind man acts). Here, just as in the case of empathetic interventions, any attempts to measure whether the target of the intervention in the case of a positive outcome (behavioral change) indeed acted in a virtuous way. A virtue of kindness may manifest itself in particular behaviors but as such cannot be identified with the virtuous deed, as highlighted by Natasza Szutta. Although again, the goal of the experiment itself was to change the behavior of the attackers, and measuring whether particular messages contributed to more flourishing individuals in terms of virtue development and character creation lies beyond the scope of this work. Interventions could affect people either way - stimulate their moral reflections or limit their attacks due to consequentialist way of thinking and fear of potential outcomes (e.g. being excluded from the community).

Additionally, Reddit created their social etiquette called "Rediquette" (https://www.reddithelp.com/hc/en-us/articles/205926439) which is "an informal expression of the values of many redditors, as written by redditors themselves." All users are encouraged to abide by it and moderators of communities (called subreddits) existing as a part of the platform have the authority to exclude its members based either on the basis of breaking the rules of the reddiquette or any other local rules imposed by specific subreddit. Volunteers were encouraged to refer to those norms as well, citing specific points, for instance, "Hey there, have you read the reddiquette? It says remember the human". or "Let's recall the reddiquette and adhere to the same standards of behavior online that you follow in real life".

As we have seen, those two categories of interventions - normative and empathetic - encompass a whole variety of categories within. One might think that it would be useful to construct a more diverse set of treatment conditions in which we test each of the ethics or empathy differentiation separately - e.g. one for virtue ethics, deontology, perhaps utilitarian approach, empathy toward the receiver of the attack, empathy towards the attacker, so on and so forth. Additionally, different types of interventions could be tested and tailored depending on the spectrum on which one can be found in the foundations of moral reasoning, as proposed by Jonathan Haidt: Care/Harm, Fairness/Cheating, Loyalty/Betrayal, Authority/Subversion, Sanctity/Degradation, and Liberty/Oppression (J. Haidt, The Righteous Mind). Albeit for the sake of the sanity of our volunteers, we decided to create only two but broad categories and let them decide based on the context of the conversation, how they want to respond to the attacker.

>>>>

---
title: "Investigating the impact of interventionist causal approach on the study of verbal aggression online"
author: "Patrycja Tempska and Rafal Urbaniak"
output:
  pdf_document:
    number_sections: yes
    df_print: kable
    keep_tex: yes
    includes:
      in_header: Rafal_latex6.sty
  html_document:
    df_print: paged
  word_document: default
classoption: dvipsnames,enabledeprecatedfontcommands
fontsize: 10pt
documentclass: scrartcl
urlcolor: blue
bibliography: ../references/attacks.bib
csl: ../references/apa-6th-edition.csl
---

@@

@@

We live in a world in which there is an abundance of data. Referring the phrase coined by Clive Humby, "data is the new oil" --- data has displaced oil and become the world’s most valuable resource. Albeit there are some fundamental differences between the two assets. Oil belongs to a very limited and finite resource, while data is growing at an exponential rate (cite: https://cloudtweaks.com/2015/03/how-much-data-is-produced-every-day/). The value of oil does not depend on its amount or the refinery processes - it's rather constant at a particular dynamic of market's supply and demand. Data and its value, on the contrary, is inherently tied to its amount and the "refinery process" - various tools and techniques for its discovery and analysis. In some cases, certain techniques can be used only if one has a sufficient (meaning enormous) data at one's disposal --- e.g. statistical methods like machine learning and deep learning \todo{@article{lecun2015deep,

@@

In the last couple of years, more and more industry and academia representatives alike started to research and develop tools for causal discovery and inference. Unfortunately, although thinking in causal terms seems to be our inherent human capacity and one of our basic cognitive abilities, that is not necessarily the case with the majority of toolkits and statistical models developed to draw insight and correlations from data. 

@@

But often scientists conduct the so-called cross-sectional research, in which no variable is manipulated. In the language and toolkit of classical statistics in such cases we cannot conduct any causal inference. What we can measure using those methods is correlation, and, as frequently repeated, “correlation does not imply causation”. Correlation can be positive, negative, or there can be no correlation at all. A positive one implies that the increase of the value of one variable will influence the increase of another variables’ value. A negative correlation means that the change in values of two variables will diverge into separate directions --- one will increase and one will decrease alongside the first one. We can measure the strength of the correlation using various mathematical methods, where 1 or -1 means we are dealing with perfect correlation --- the value of one variable changes proportional to the value of another variable. But even though causation is necessarily connected with correlation, as already mentioned --- not every correlation implies causation. This is especially problematic if we take into consideration the fact that statisticians and people in general naturally think in causal terms and spurious interpretations of studies occur frequently. On the other hand, a conservative approach of many classical statisticians led to a long-living debate on whether smoking causes cancer, lasting for over 20 years. One of the reasons for such a fierce debate was the lack of a causal framework, which was still in its infancy at that time. Framework of tools with which we could validate or invalidate assumptions about the existence of the potential confounder --- a variable causing both craving cigarettes and lung cancer (e.g. some genetic factor), creating a spurious correlation between cigarettes and lung cancer.
Modeling causality and moving beyond traditional statistics is a relatively young endeavor. One of the pioneers of the work on causality is Judea Pearl. As he describes, causality and traditional statistics are governed by opposing ways of thinking. While statistics might be described as “study of methods of the reduction of data”, causality requires the incorporation of an extra layer --- the understanding or prior knowledge (assumptions) about the process that generate the data in the first place. Classical statistics might be described here as a “model-free approach,” just as Artificial Intelligence methods based on machine learning/deep learning, in which no prior knowledge can be taken into account.

In this paper, I will show how mathematical methods for process modeling which includes assumptions about causal connections can be applied to discover how different variables affect each other, moving beyond classical statistics. Looking at the analysis from two perspectives --- traditional (classical, frequentist statistics) and bayesian -- we will see the advantage of the second paradigm in learning about the world based on the data we have at our disposal. Data used for analysis concerned in this paper comes from an experimental intervention study conducted in a naturalistic, digital setting (Q&A forum on Reddit), utilizing a collective intelligence approach to content moderation and reduction of the level of verbal aggression among a selected group of Reddit users who regularly attack other community members. Collective Intelligence in this sense means exploring the collaboration between human and machine intelligence to develop solutions to social challenges. Artificial Intelligence was used to detect verbal aggression (personal attacks) and notify human volunteers about attacks. Volunteers after receiving notifications employed interventions inspired in its underlying principles in philosophy and psychology. There were two broad categories of interventions - one based on norms (deontology, virtue ethics among others) and the second based on empathy (inspired by notions of David Hume and Adam Smith).

One might also wonder about the motivations for conducting such an experiment. Much effort has been made in order to tackle the problem of verbal aggression and harassment online, but looking at various reports and surveys [@sorrentino2019epidemiology; @vogels_state_2021; @Zachary_hate], it remains a common hindrance for people engaging with social media in their everyday lives. The problem got exacerbated in the midst of the COVID19 pandemic, during which the majority of our social life moved to cyberspace. During this shift, there was an increase in cyberbullying attitudes and perpetration [@barlett2021comparing], 90% increase in public reports of illegal online content [@grant_2021], including 114% increase in non-consensual sharing of intimate images, 30% increase in cyberbullying, as well as 40% of increase in adults reporting online harassment. According to a report conducted by company L1ght [@noauthor_l1ght_2020], hate speech directed towards China and the Chinese went up by 900% on Twitter. Gaming platforms were in the spotlight as well, with a 40% increase in toxicity on Discord \footnote{In both reports, the increase is reported as a relative change between the year 2019 and 2020, with no absolute indicators.}.

But alongside the growing need for even more efficient and proactive moderation, the capacity to execute it did not go hand in hand, forcing companies and policymakers to rethink the current model of moderation processes and workforce. Due to the COVID19 restrictions including social distancing, a lot of those serving the role of moderators had to be sent home [@bhattacharya] without the ability to work remotely because of the constraints affiliated with restrictive non-disclosure agreements (NDA) among others. Curtailing the moderators' workforce was accompanied by more agency given to algorithms and AI-based moderation. Those changes, as argued by @gerrard2020covid19, can be seen as a serious red flag in terms of safety for all users on online platforms. That is why in the paper I will also try to sketch a landscape of current moderation techniques through a critical lens.

@@

# Collective Intelligence Approach to Counter-speech
Traditionally collective intelligence has been defined as "a group or a team’s combined capacity and capability to perform a wide variety of tasks and solve diverse problems" [@noauthor_collective_nodate]. In our paper and in the theoretical underpinnings of the experiment itself, we will be relying on a collective intelligence scope proposed by Nesta, an innovation foundation (https://www.nesta.org.uk), which focuses on a collaboration between human and machine intelligence to develop innovative solutions to social challenges.

@@

\textbf{Empathetic interventions}
In the first treatment condition, volunteers were asked to send empathy-inducing messages focusing either on the target of verbal aggression (e.g. „Hey such words might hurt the other person”), stressing the common humanity aspect that we call share („We are all humans of flesh and blood”), or even infusing the intervention with the emphatic response to the attacker (Hey I understand your strong emotions…”). 
At the core of the empathetic interventions, we put forth the notion that goes back to David Hume and Adam Smith. The first one conceived empathy (at the time referred to as sympathy) as mirroring the emotional state of another person. In the academic psychological literature, similar phenomenon was distinguished and coined in the term emotional contagion
(https://www.sciencedirect.com/topics/psychology/emotional-contagion): "the process in which an observed behavioral change in one individual leads to the reflexive production of the same behavior by other individuals in close proximity, with the likely outcome of converging emotionally".

For Adam Smith, sympathy consisted of visualizing how the sympathetic person would feel in the particular circumstances of the other --- thus here the process was based not so much on mirroring, but rather projecting my imagination of what it is like to be that person in a certain moment). Without further delving into the differentiation between those two, sympathy in both accounts is crucial in the constitution of human beings as social and moral creatures (https://plato.stanford.edu/entries/empathy/). It enables the emotional connection to others and concern for their well-being. The interventions were supposed to serve the role of empathetic response enablers --- reminding the person in front of the screen that there is a real human being on the other side.

In the psychological literature, various kinds of empathetic responses were distinguished - the aforementioned emotional contagion, affective/proper empathy, sympathy, personal distress or cognitive empathy ((https://plato.stanford.edu/entries/empathy/). During the experiment, we could not observe or measure whether the empathetic response indeed was evoked. The study was conducted in a digital setting --- interventions were most likely read in the private space, in front of personal computers or smartphones. Even if we would change the setting to a lab, differentiating and measuring empathy is not trivial if not impossible (to check).

Also, interventions that stated that "such words might hurt the other person" have the underlying assumption that the receiver of the attack might be hurt, but in reality that might not even be the case. Intuitively, and following the argumentation of Thomas Nagel in “What is it like to be a bat?” one can only imagine what it is like --- but for me --- to be a bat or to be a receiver of the attack. To each observed or imagined experience there is an array of subjective quality to actually experiencing it and in this way receiving a particular message might be met with a unique reception by each conscious being. Also, full epistemic access to the mind and body of another is impossible.

But even though such access is impossible and putting aside the broad spectrum of phenomena related to empathetic response, our goal was by utilizing interventions referring to empathy in various forms, changing the behavior of the attacker --- convincing him to refrain from using verbal aggression towards others. Whether the message indeed gave rise to empathy was not of importance --- one can imagine a hypothetical scenario of a successful intervention in which the attacker changed his behavior long-term and stopped attacking others, hopefully as a result of interventions, but the empathetic response was not even evoked. The behavioral change could be induced by other motivations --- e.g. unwillingness from the exclusion from the community. Thus the goal is not to evoke empathy, but through empathetic interventions --- substantially limit the use of verbal aggression among study group members.

\textbf{Normative interventions}
In the second treatment condition, volunteers were asked to send norms-inducing messages written on the grounds of deontology (e.g. „we have a duty to respect each other during the discussions"), virtue ethics (e.g. „capacity to be respectful in a heated discussion is a virtue and requires hard work”) or Reddit's community guideline (e.g. "Let's recall the reddiquette and adhere to the same standards of behavior online that you follow in real life").

Whatever the strategy was used, it was supposed to express the unacceptability of verbal aggression in a direct or non-direct way. A more direct approach referred to deontology in which „actions are good or bad according to a clear set of rules” (https://ethics.org.au/ethics-explainer-deontology/). The basic assumption behind those interventions is that there is something we ought to do or are morally required to do --- and such actions are the right actions. There are also actions we ought not to do --- and such actions are morally wrongful. This line of thinking can be traced back to Immanuel Kant who thought that all universal moral obligations can stem from categorical imperative: “act only in accordance with that maxim through which you can at the same time will that it become a universal law” (Groundwork of the Metaphysic of Morals).

In the psychological literature, Robert B. Cialdini distinguished three types of norms that he found to be effective in guiding human action: descriptive norms (focused on the perception of how the majority of people would behave), injunctive (based on the perception of how the majority of people would approve or disapprove certain conduct), or personal norms (based on the perception of how a particular person would approve or disapprove certain behavior) CITE: https://www.sciencedirect.com/science/article/abs/pii/S0065260108603305. Although interventions written on the grounds of duties and deontology cannot be reduced to any of the above, since following Hume's guillotine --- from the fact that the majority of people are respectful does not follow the normative statement of civility --- I hoped that expressing a universal normative statement will be even more powerful and potentially spawn an interpretation of social acceptability or unacceptability of certain behavior.

Yet other interventions in normative group were encouraged to be expressed in the light of virtue ethics and stressed the importance of adhering to or practicing certain virtues. Such a message was assumed to motivate the moral agent in the process of positive reinforcement and create a feeling of the desirability of certain behaviors, like kindness and civility. The core assumption here is that the virtue itself is not genetically inherited but is rather a potential or a disposition of a character that can be practice and mastered like a practical skill (even though as mentioned by Natasza Szutta in [cite her book about virtue ethics] virtues and practical skills share some fundamental differences). NS distinguished two types of virtues - affective and cognitive. Affective virtues encompass the emotions and feelings that play an important and positive role in morality and can act as a support in the course of becoming a virtuous man. A cognitive virtue relates to the intellectual aspect in which one knows how to act in certain situations and understands the rules of morally rightful actions (e.g. what is kindness and how a kind man acts).

Here, just as in the case of empathetic interventions, any attempts to measure whether the target of the intervention in the case of a positive outcome (behavioral change) indeed acted in a virtuous way. A virtue of kindness may manifest itself in particular behaviors but as such cannot be identified with the virtuous deed, as highlighted by Natasza Szutta. Although again, the goal of the experiment itself was to change the behavior of the attackers, and measuring whether particular messages contributed to more flourishing individuals in terms of virtue development and character creation lies beyond the scope of this work. Interventions could affect people either way - stimulate their moral reflections or limit their attacks due to consequentialist way of thinking and fear of potential outcomes (e.g. being excluded from the community).

Additionally, Reddit created their social etiquette called "Rediquette" (https://www.reddithelp.com/hc/en-us/articles/205926439) which is "an informal expression of the values of many redditors, as written by redditors themselves." All users are encouraged to abide by it and moderators of communities (called subreddits) existing as a part of the platform have the authority to exclude its members based either on the basis of breaking the rules of the reddiquette or any other local rules imposed by specific subreddit. Volunteers were encouraged to refer to those norms as well, citing specific points, for instance, "Hey there, have you read the reddiquette? It says remember the human".

As we have seen, those two categories of interventions - normative and empathetic - encompass a whole variety of categories within. One might think that it would be useful to construct a more diverse set of treatment conditions in which we test each of the ethics or empathy differentiation separately - e.g. one for virtue ethics, deontology, perhaps utilitarian approach, empathy toward the receiver of the attack, empathy towards the attacker, so on and so forth. Additionally, different types of interventions could be tested and tailored depending on the spectrum on which one can be found in the foundations of moral reasoning, as proposed by Jonathan Haidt: Care/Harm, Fairness/Cheating, Loyalty/Betrayal, Authority/Subversion, Sanctity/Degradation, and Liberty/Oppression (J. Haidt, The Righteous Mind). Albeit for the sake of the sanity of our volunteers, we decided to create only two but broad categories and let volunteers decide based on the context of the conversation, how they want to respond to the attacker.
